---
title: "Parametric vs. Non-Parametric Tests"
author: "Hector Mathonsi"
date: "13 October 2024"
output:
    html_notebook:
        toc: true
        toc_float: true
---

# Data Sets
```{r}
# Create the following vectors/data
data1 <- c(84,86,90,97,98,98,99,102,102,103,108,111)
data2 <- c(91,95,96,96,96,97,98,101,103,105,106,123,130,143,145)
test_data <- cbind(1:8,c(59,68,60,60,59,62,58,76),c(61,78,65,56,55,62,60,68))
colnames(test_data) <- c("Student","Test1","Test2")
test_data <- as.data.frame(test_data)
test_data

# Create the growth dataset
# A researcher wants to know if three different fertilizers lead to different levels of plant growth. A sample of 30
# plants are randomly divided into three groups of 10 and a different fertilizer applied to each group. The height of
# each plant was measured at the end of one month.
growth <- data.frame(group=rep(c('A', 'B', 'C'), each=10), height=c(7, 14, 14, 13, 12, 9, 6, 14, 12, 8, 15, 17, 13, 15, 15, 13, 9, 12, 10, 8, 6, 8, 8, 9, 5, 14, 13, 8, 10, 9))
growth

# The cars dataset is available in R
# The data give the speed of 50 cars and the distances taken to stop.
# speed - Speed (mph)
# dist - Stopping distance (ft)
data(cars)
cars

# The mtcars dataset is available in R
# The data give fuel consumption and 10 aspects of automobile design and performance for 32 cars.
data(mtcars)
mtcars
```

# One-Sample Test
The lecturer wants to test if the average/median mark for test 1 is significantly better than the average/median mark of last year, namely: mean = 56 and median = 55.
```{r}
# Test normality
shapiro.test(test_data$Test1)

# One-sample test - t-test (If the data is normally distributed)
t.test(test_data$Test1, mu = 56, alternative = "greater")

# One-sample test - wilcox-test (If the data is not normally distributed)
wilcox.test(test_data$Test1, mu = 55, alternative = "greater", exact = FALSE)
```

#  Paired sample test
The lecturer wants to know if there is a significant difference in students’ marks for test 1 compared to test 2.
```{r}
# Test normality
shapiro.test(test_data$Test1)
shapiro.test(test_data$Test2)

# One-sample test - t-test (If the data is normally distributed)
t.test(test_data$Test1, test_data$Test2, paired = T)

# One-sample test - wilcox-test (If the data is not normally distributed)
wilcox.test(test_data$Test1, test_data$Test2, paired = T, exact = FALSE)
```

# Two Independent Sample Tests
```{r}
summary(data1)
summary(data2)

# Check assumptions
shapiro.test(data1)
shapiro.test(data2)

# Check Variance
var.test(data1, data2)


# Independent samples t-test
t.test(data1, data2, var.equal = FALSE)

# Mann-Whitney U test
wilcox.test(data1, data2, exact = FALSE)
```

# Multiple Independent Sample Tests
```{r}
# Three independent samples (growth data)
# Box plot
boxplot(height ~ group, data = growth)

# Check Assumptions
shapiro.test(growth$height[growth$group == "A"])
shapiro.test(growth$height[growth$group == "B"])
shapiro.test(growth$height[growth$group == "C"])

bartlett.test(height ~ group, data = growth)    # Equal variance test for normal data
fligner.test(height ~ group, data = growth)     # Equal variance test for non-normal data

# One-way ANOVA
aov(height ~ group, data = growth)

# Kruskal-Wallis test
kruskal.test(height ~ group, data = growth)
```

# Correlation Tests
```{r}
plot(cars$speed, cars$dist)
cor.test(cars$speed, cars$dist, method = "pearson")
cor.test(cars$speed, cars$dist, method = "spearman")
cor.test(cars$speed, cars$dist, method = "kendall")
```





