---
title: "Chi-Squared Goodness-of-Fit Test"
author: "Hector Vukosi Mathonsi"
date: "`r Sys.Date()`"
output: html_document
---

## Introduction

The Chi-Squared Goodness-of-Fit Test is used to determine if an observed categorical data distribution matches an expected distribution. In this example, we'll demonstrate how to perform this test using R.

## Step 1: Define Observed and Expected Frequencies

First, define the observed and expected frequencies of the categories you are analyzing.

```{r}
# Observed frequencies
observed <- c(50, 30, 20)

# Expected frequencies
expected <- c(40, 30, 30)

# Display the frequencies
data.frame(Observed = observed, Expected = expected)
```

## Step 2: Perform the Chi-Squared Test

You can use the `chisq.test()` function to perform the Chi-Squared Goodness-of-Fit Test. Specify the observed frequencies and the expected probabilities (expected frequencies divided by the total observations).

```{r}
# Perform the Chi-Squared Goodness-of-Fit Test
test_result <- chisq.test(observed, p = expected / sum(expected))

# Display the result
test_result
```

## Step 3: Interpret the Results

The key output from the Chi-Squared test is the test statistic and the p-value. Based on the p-value, you can decide whether to reject or fail to reject the null hypothesis.

- **Null Hypothesis (H0)**: The observed data fits the expected distribution.
- **Alternative Hypothesis (H1)**: The observed data does not fit the expected distribution.

The decision rule is:
- If the p-value is less than the significance level (commonly 0.05), reject the null hypothesis.
- If the p-value is greater than or equal to the significance level, fail to reject the null hypothesis.

```{r}
# Check p-value
if (test_result$p.value < 0.05) {
  cat("We reject the null hypothesis: The observed distribution differs from the expected distribution.")
} else {
  cat("We fail to reject the null hypothesis: The observed distribution fits the expected distribution.")
}
```

## Step 4: Visualize the Data (Optional)

You can also visualize the observed and expected frequencies to gain a better understanding of how they compare.

```{r}
# Create a bar plot to compare observed and expected frequencies
barplot(rbind(observed, expected), beside = TRUE, col = c("skyblue", "lightgreen"),
        legend.text = c("Observed", "Expected"),
        main = "Comparison of Observed and Expected Frequencies",
        xlab = "Categories", ylab = "Frequency")
```

## Conclusion

In this example, we performed the Chi-Squared Goodness-of-Fit Test and used the p-value to interpret whether the observed data matches the expected distribution. Based on this test, we can make conclusions about the fit of our data.
```
